Code Club Project: Assessing whether intra and inter-genomic variation hinder utility of ESVs and ASVs

### 2021-01-11 submission/manuscript.Rmd
Insert code for first paragraph of results

---
output:
  word_document: default
  pdf_document:
    keep_tex: true
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
---

@@

To investigate the variation in the number of copies of the 16S rRNA gene per genome as well as the intragenomic variation among copies of the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn copy number database (rrnDB; CITATION). Among the XX,XXX genomes represented in the rrnDB, each genome had between X (e.g. XXXXXXX) and XX (e.g. XXXXXXX) copies of the rrn operon. As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased (**FIGURE**). On average, there were XXX variants per copy of the full length 16S rRNA gene and an average of XXX, XXX, and XXX variants when considering the V4, V3-V4, and V4-V5 regions of the gene. Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**FIGURE**). For example, M. tuberculosis generally only had one copy of the gene per genome, but across the XXX genomes that have been sequenced there were XX versions of the gene. Similarly, a E. coli genome typically had 7 copies of the 16S rRNA gene with between X and X distinct full length sequences per genome. Across the XXX genomes that have been sequenced, there were XXX different variants of the gene. These observations highlight the risk of selecting a threshold for defining units of inference that is too narrow because it is possible to split a single genome into multiple units.

>>>>

---
output:
  word_document: default
  pdf_document:
    keep_tex: true
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE}
library(tidyverse)
library(knitr)
library(here)
```

@@

```{r, echo=FALSE}
metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
										 col_types = cols(.default = col_character()))

easv <- read_tsv(here("data/processed/rrnDB.easv.count_tibble"),
								col_types = cols(.default = col_character(),
																 count = col_integer()))

metadata_easv <- inner_join(metadata, easv, by=c("genome_id" = "genome")) %>%
	mutate(threshold = recode(threshold, "esv" = "0.000"),
				 threshold = as.numeric(threshold))

rrn_copy_number <- metadata_easv %>% 
	filter(region == "v19", threshold == 0) %>%
	group_by(species, genome_id) %>%
	summarize(n_rrn = sum(count), .groups="drop") %>%
	group_by(species) %>%
	summarize(min = min(n_rrn), max = max(n_rrn), median = median(n_rrn), N=n(), .groups="drop")

single_copy <- rrn_copy_number %>% filter(median == 1) %>% top_n(n=1, N)
max_copy <- rrn_copy_number %>% top_n(n=1, median)

single_copy_N <- rrn_copy_number %>% filter(species == single_copy$species) %>% pull(N)
max_copy_N <- rrn_copy_number %>% filter(species == max_copy$species) %>% pull(N)
ecoli_copy <- rrn_copy_number %>% filter(species == "Escherichia coli")

rates <- metadata_easv %>%
	filter(threshold == 0) %>%
	group_by(region, genome_id) %>%
	summarize(n_rrns = sum(count), n_asvs = n_distinct(easv), .groups="drop") %>%
	group_by(region) %>%
	summarize(rate = sum(n_asvs)/sum(n_rrns), .groups="drop")

rate_v19 <- rates %>% filter(region == "v19") %>% pull(rate)
rate_v4 <- rates %>% filter(region == "v4") %>% pull(rate)
rate_v34 <- rates %>% filter(region == "v34") %>% pull(rate)
rate_v45 <- rates %>% filter(region == "v45") %>% pull(rate)

asvs_per_species <- metadata_easv %>%
	filter(threshold == 0 & region == "v19") %>%
	group_by(species) %>%
	summarize(n_asvs = n_distinct(easv), .groups="drop")

single_nasvs <- asvs_per_species %>% filter(species == single_copy$species) %>% pull(n_asvs)
ecoli_nasvs <- asvs_per_species %>% filter(species == "Escherichia coli") %>% pull(n_asvs)
```

To investigate the variation in the number of copies of the 16S rRNA gene per genome as well as the intragenomic variation among copies of the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn copy number database (rrnDB; CITATION). Among the `r n_distinct(metadata$species)` species represented in the rrnDB there were `r nrow(metadata)` genomes. The median number of rrn operson per species ranged between `r single_copy$median` (e.g. `r single_copy$species`) and `r max_copy$median` (e.g. `r max_copy$species`) copies of the rrn operon. As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased (**FIGURE**). On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v4`, `r rate_v34`, and `r rate_v45` variants when considering the V4, V3-V4, and V4-V5 regions of the gene, respectively. Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**FIGURE**). For example, `r single_copy$species` generally only had `r single_copy$median` copy of the gene per genome, but across the `r single_copy_N` genomes that have been sequenced there were `r single_nasvs` versions of the gene. Similarly, a E. coli genome typically had `r ecoli_copy$median` copies of the 16S rRNA gene with between `r ecoli_copy$min` and `r ecoli_copy$max` distinct full length sequences per genome. Across the `r ecoli_copy$N` E. coli genomes that have been sequenced, there were `r ecoli_nasvs` different variants of the gene. These observations highlight the risk of selecting a threshold for defining units of inference that is too narrow because it is possible to split a single genome into multiple units.

### 2021-01-11 submission/manuscript.tex
Insert code for first paragraph of results

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

\begin{itemize}
\item
  16S rRNA gene sequencing is a very powerful technique for describing
  and comparing microbial communities
\item
  How do we analyze them (classification, clustering)?
\item
  What has changed in recent years? ASVs
\item
  Efforts to link 16S rRNA gene sequences to taxonomic levels based on
  distance thresholds go back a long way
\item
  ESVs/ASVs have been an attempt to adopt the thresholds suggested by
  genome sequencing to microbial community analysis using 16S rRNA gene
  sequences
\item
  Most bacterial genomes have more than 1 copy of the rrn operon and
  those copies are not identical
\item
  Using too fine a threshold to create taxonomic groups runs risk of
  splitting single genome into multiple bins
\item
  For example, E. coli K-12 has 7 copies of the 16S rRNA gene with 5
  variants
\item
  Using too broad a threshold to define ASVs or OTUs risks lumping
  together bacterial species into the same grouping
\item
  For example, B. cereus, thuringiensis, anthracis share the same 16S
  rRNA gene sequences
\item
  Goal of this study
\end{itemize}

@@

\hypertarget{results}{%
\subsection{Results}\label{results}}

\begin{itemize}
\tightlist
\item
  ESVs/ASVs

  \begin{itemize}
  \tightlist
  \item
    copy number varies by taxonomy
  \item
    more copies, more variants per genome
  \item
    full length sequences have more variants than sub-regions
  \item
    as more sequences are added to a species, the number of variants
    increases
  \end{itemize}
\item
  OTUs

  \begin{itemize}
  \tightlist
  \item
    increasing a threshold decreases the number of variants
  \item
    this limits the splitting of a single genome into multiple bins
  \item
    this increases the lumping of species into single bin
  \end{itemize}
\end{itemize}

\hypertarget{conclusions}{%
\subsection{Conclusions}\label{conclusions}}

\begin{itemize}
\item
  Briefly synthesize results

  \begin{itemize}
  \tightlist
  \item
    Unlikely that the unit of inference should be an ASV
  \end{itemize}
\item
  No biological argument to split a genome into multiple bins
\item
  This analysis has allowed some splitting to balance with lumping
\item
  To reduce splitting further, you would need larger thresholds
\item
  There is general agreement in the field that if you want to classify
  something to a bacterial species, you need more than the 16S rRNA gene
\item
  Furthermore, using only a few hundred bases of that gene are even more
  limited.
\item
  We are asking too much of a short section of sequence
\item
  Surprisingly, 3\% performs pretty well for an operational definition
  that limits splitting of bacterial genomes and avoiding the lumping of
  bacterial species
\end{itemize}

\hypertarget{materials-and-methods}{%
\subsection{Materials and Methods}\label{materials-and-methods}}

\begin{itemize}
\tightlist
\item
  rrnDB
\item
  NCBI taxonomy
\item
  R and R packages
\item
  GitHub / YouTube
\end{itemize}

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

>>>>

16S rRNA gene sequencing is a very powerful technique for describing and
comparing microbial communities. Efforts to link 16S rRNA gene sequences
to taxonomic levels based on distance thresholds go back to at least the
1990s. The distance-based thresholds that were developed and are now
widely used (3\%) were based on DNA-DNA hybridization approaches that
are not as precise as genome sequencing. Instead, genome sequencing
technologies have suggested that the widely used 3\% distance threshold
to operationally define bacterial taxa is too coarse. As an alternative
to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way
to adopt the thresholds suggested by genome sequencing to microbial
community analysis using 16S rRNA gene sequences. ASVs are a unit of
microbial community inference that do not cluster sequences based on a
distance-based threshold. However, most bacterial genomes have more than
1 copy of the rrn operon and those copies are not identical. Therefore,
using too fine a threshold to identify OTUs creates the risk of
splitting a single genome into multiple bins and using too broad of a
threshold to define OTUs creates the risk of lumping together bacterial
species into the same OTU. An example of both is seen in the comparison
of Staphylococcus aureus (NCTC 8325) and S. epidermidis (ATCC 12228)
where each genome has 5 copies of the 16S rRNA gene. The 10 copies of
the 16S rRNA gene each have a different sequence and so if OTUs are
defined based on ASVs, each genome would be split into 5 OTUs.
Conversely, if the copies were clustered using a 3\% distance threshold
all 10 copies would cluster into the same OTU. The goal of this study
was to quantify the risk of splitting a single genome into multiple bins
and the risk of lumping together different bacterial species into the
same bin.

To investigate the variation in the number of copies of the 16S rRNA
gene per genome as well as the intragenomic variation among copies of
the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn
copy number database (rrnDB; CITATION). Among the 4774 species
represented in the rrnDB there were 15614 genomes. The median number of
rrn operson per species ranged between 1 (e.g.~Mycobacterium
tuberculosis) and 19 (e.g.~Metabacillus litoralis) copies of the rrn
operon. As the number of copies of the operon in a genome increased, the
number of variants of the 16S rRNA gene in each genome also increased
(\textbf{FIGURE}). On average, there were 0.5970303 variants per copy of
the full length 16S rRNA gene and an average of 0.2551013, 0.3265188,
and 0.2744228 variants when considering the V4, V3-V4, and V4-V5 regions
of the gene, respectively. Although a species tended to have a
consistent number of 16S rRNA gene copies per genome, the number of
total variants increased with the number of genomes that were sampled
(\textbf{FIGURE}). For example, Mycobacterium tuberculosis generally
only had 1 copy of the gene per genome, but across the 180 genomes that
have been sequenced there were 11 versions of the gene. Similarly, a E.
coli genome typically had 7 copies of the 16S rRNA gene with between 6
and 10 distinct full length sequences per genome. Across the 958 E. coli
genomes that have been sequenced, there were 1013 different variants of
the gene. These observations highlight the risk of selecting a threshold
for defining units of inference that is too narrow because it is
possible to split a single genome into multiple units.

A method to avoid splitting a single genome into multiple units of
inference is to cluster 16S rRNA gene sequences together that are
similar. However, this also increases the risk of lumping together genes
from different species that are similar to each other. Therefore, I
assessed the impact of the threshold used to define clusters of 16S rRNA
genes on the propopensity to lump species together and split genome
apart. I identified the threshold where 90\% of bacterial species would
be represented by a single OTU. For full length 16S rRNA gene sequences,
I found that at a threshold of XX\%, 90\% of the species would be
represented by a single OTU. Similarly, thresholds of XX, XX, and XX\%
were observed for the V4, V3-V4, and V4-V5 regions. However, at these
thresholds, multiple species could be represented by the same OTU. At
the highest level of resolution, XX\% of the species shared a 16S rRNA
gene sequence variant with another species. Given the risk of splitting
a genome into multiple OTUs is more biologically problematic than
lumping species together, larger thresholds are advisable.

To provide a more nuanced approach to selecting a threshold, it would be
useful to to quantify the sensitivity and specificity of characterizing
bacterial species using OTUs defined at different thresholds. I created
confusion matrices for multiple regions of the 16S rRNA gene: true
positives were those cases where two ASVs were joined in the same OTU
and the same species; true negatives were those cases where two ASVs
from different OTUs came from different species; false positives were
those ASVs that joined the same OTU, but were from different species;
and false negatives were those ASVs that joined different OTUs, but were
from the same species. By calculating the sensitivity and specificity
for each threshold and each region of the 16S rRNA gene, I was able to
constuct a receiver operator characteristic curve (ROC). Because the ROC
curve represents a range of possible thresholds and sensitivities and
specificities, I used two metrics to select the best threshold for
defining an OTU. First, I identified the thresholds where the
sensitivity and specificity were most similar to each other. For this
criterion, the best thresholds were XX (V1-V9), XX (V4), XX (V3V4), and
XX (V4V5). Second, I identified the threshold that resulted in the point
on the ROC curve that was closest to perfect classification. For this
criterion, the best thresholds were XX (V1-V9), XX (V4), XX (V3V4), and
XX (V4V5). Surprisingly, these analyses revealed that thresholds near
3\% balance the risks of splitting genomes into separate OTUs and
lumping species into the same OTU.

The results of this analysis demonstrate that there is a significant
risk of splitting single genomes into multiple bins if too fine of a
threshold is applied to defining an OTU. An ongoing problem for
amplicon-based studies is defining a meaningful taxonomic unit of
inference. Since there is no consensus definition for a biological
species concept, microbiologists must accept that how we have named
bacterial species is biased and that taxonomic rules are not applied in
a consistent manner. This makes it more challenging to attempt to fit a
distance threshold to define an OTU definition that matches a set of
species names. Furthermore, it is unlikely that the 16S rRNA gene
evolves at the same rate across all bacterial lineages, which limits the
biological interpretation of a common OTU definition. At best, a
distance-based definition of a taxonomic unit is operational. There is
general agreement in bacterial systematics that to classify something to
a bacterial species, you need phenotypic and genome sequence data
(CITATION). We are asking too much of a short section of a bacterial
genome to be able to differentiate between species. It is difficult to
defend a unit of inference that would split a single genome into
multiple taxonomic units. It is not biologically plausible to entertain
the possability that parts of a genome would have different ecologies.
Although there are multiple reasons that proponents of ASVs encourage
their use, the significant risk of splitting genomes is too high to
warrant their use.

\textbf{Materials and Methods. (i) Data availability.} The 16S rRNA gene
sequences used in this study were obtained from the \emph{rrn}DB
(\url{https://rrndb.umms.med.umich.edu}; version 5.6, released November
8, 2019). At the time of submission, this is the most current version of
the database. The \emph{rrn}DB obtained the curated 16S rRNA gene
sequences from the KEGG database, which ultimately obtained them from
NCBI's non-redundant RefSeq database. The \emph{rrn}DB provides
downloadable versions of the sequences with their taxonomy as determined
using the naive Bayesian classifier trained on the RDP reference
taxonomy. For some genomes this resulted in multiple classifications
since a genome's 16S rRNA gene sequences were not identical. Instead, I
mapped the RefSeq accession number for each genome in the database to
obtain a single taxonomy for each genome. Because strain names were not
consistently given to genomes across bacterial species, the strain level
designations were ignored.

\textbf{(ii) Definition of regions within 16S rRNA gene.} The full
length 16S rRNA gene sequences were aligned to a SILVA reference
alignment of the 16S rRNA gene (v138) using the mothur software package
(v. 1.XX). Regions of the 16S rRNA gene were selected because of their
use in the microbial ecology literature. Full length sequences
corresponded to \emph{E. coli} positions XX through XXXX, V4 to
positions XXX through XXX, V3-V4 to positions XXX through XXX, and V4-V5
to positions XXX through XXX.

\textbf{(iii) Controlling for uneven sampling of genomes by species.}
Because of the uneven distribution of genome sequences across species,
for the analysis of splitting genomes and lumping species I randomly
selected one genome for each species. The random selection was repeated
100 times. Analyses based on this randomization report the median of the
100 randomizations. The intraquartile range between randomizations was
typically less than XXXX. Because it was so small, confidence intervals
are not included in Figure 2.

\textbf{(iv) Reproducible data analysis.} The code to perform the
analysis in this manuscript and its hisotry are available as a git-based
version control repository on GitHub
(\url{https://github.com/pschloss/Schloss_rrnAnalysis_XXXX_2020}). The
analysis can be regenerated using a GNU Make-based workflow that made
use of built-in bash tools (v. 3.2.57), mothur (v. 1.XX), and R (v.
4.X.X). Within R, I used the tidyverse (v. 4.X.X), data.table (v.
4.X.X), Rcpp (v. 4.X.X), furrr (v. 4.X.X), and rmarkdown (v. 4.X.X)
packages. The conception and development of this analysis is available
as a playlist on the Riffomonas YouTube channel
(\url{https://www.youtube.com/playlist?list=PLmNrK_nkqBpKY3SZiivlIGvcLX-KHmfR8}).

\textbf{Acknowledgements.} I am grateful to Robert Hein and Thomas
Schmidt who maintain the rrnDB for their help in understanding the
curation of the database and for making the 16S rRNA gene sequences and
related metadata publicly available. I am also grateful to community
members who watched the serialized version of this analysis on YouTube
and provided their suggestions and questions.

This work was supported, in part, through grants from the NIH to PDS
(P30DK034933, U01AI124255, and R01CA215574).

@@

\hypertarget{references}{%
\subsection{References}\label{references}}
